How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-05-27 talk/dyla2007/dyla.tex
spell check and small remaining fixes

In this paper, we will argue that it is possible to
benefit from and integrate with OO VMs while keeping the dynamic
language implemented with a single, simple source code base.  The idea is
to write an interpreter for that language in another sufficiently
high-level but less dynamic language.  This interpreter plays the role
of a specification for the dynamic language.  With a sufficiently capable
translation toolchain we can then generate whole virtual machines from
this specification -- either wholely custom VMs for C-level operating
systems or VMs layered on top of various OO VMs.  In other words,
meta-programming techniques can be used to successfully replace a
foreseeable one-VM-fits-all standardization attempt.

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' an
interpreter playing the role of a high-level description of the language.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe --
mostly in term of flexibility.  Most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping several of them in sync.  The single
source can be used to generate either custom VMs for C-like
environments or interpreters running on top of OO VMs.  This makes it
practical to experiment with large changes to the language and with
entirely new languages, such as domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/Posix to the JVM to .NET.

@@

We implemented this idea in the PyPy project \cite{pypy}.  The dynamic language
for which we wrote an interpreter is Python.  It is a language which,
because of its size and rather intricate semantics, is a good target for
our approach in the following sense: its previous reimplementations
(Jython for the JVM and IronPython for .NET) have each proved to be very
time-consuming to maintain.  Our implementation is, by construction,
easier to maintain and extremely portable (including to C/Posix, to the
JVM and to .NET).

@@

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\cite{LLVM} and .NET are described in \cite{architecture} \cite{translationdoc}
These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.  These figures do not
include the spectacular speed-ups obtained in some cases by the JIT compiler
described in section \ref{subsect:dynamic_compilers}.

@@

%XXX doesn't look entirely nice
\begin{itemize}
\item \emph{Do not write dynamic language implementations ``by hand''.}
Writing them more abstractly, at a higher level, has primarily only
advantages, among them the avoidance of a proliferation of implementations
growing out of sync.  Writing interpreters both flexibly and efficiently
is difficult and meta-programming is a good way to achieve it.
Moreover, this is not incompatible with targetting and benefiting from
existing high-quality object-oriented virtual machines like those of the Java and .NET.

@@

The approach we outlined is actually just one in a very large, mostly
unexplored design space; it is likely that some of the choices made in
PyPy will turn out to be suboptimal.  We are hoping that other
toolchains will emerge over time, exploring other aspects and proposing
other solutions.  By their ``meta'' nature, these multiple approachs
should be easier to bridge together than, say, multiple OO VMs with
different object and runtime models.  We believe that further research
in this area might open the door to better solutions for
interoperability in general -- e.g. high-level bridges instead of
(virtual-)machine-level ones, enabled by cross-translation.

>>>>

In this paper, we will argue that it is possible to
benefit from and integrate with OO VMs while keeping the dynamic
language implemented with a single, simple source code base.  The idea is
to write an interpreter for that language in another sufficiently
high-level but less dynamic language.  This interpreter plays the role
of a specification for the dynamic language.  With a sufficiently capable
translation toolchain we can then generate whole virtual machines from
this specification -- either wholly custom VMs for C-level operating
systems or VMs layered on top of various OO VMs.  In other words,
meta-programming techniques can be used to successfully replace a
foreseeable one-VM-fits-all standardization attempt.

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' an
interpreter playing the role of a high-level description of the language.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe --
mostly in term of flexibility.  Most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping several of them in sync.  The single
source can be used to generate either custom VMs for C-like
environments or interpreters running on top of OO VMs.  This makes it
practical to experiment with large changes to the language and with
entirely new languages, such as domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/POSIX to the JVM to .NET.

@@

We implemented this idea in the PyPy project \cite{pypy}.  The dynamic language
for which we wrote an interpreter is Python.  It is a language which,
because of its size and rather intricate semantics, is a good target for
our approach in the following sense: its previous reimplementations
(Jython for the JVM and IronPython for .NET) have each proved to be very
time-consuming to maintain.  Our implementation is, by construction,
easier to maintain and extremely portable (including to C/POSIX, to the
JVM and to .NET).

@@

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\cite{LLVM} and .NET are described in \cite{architecture} \cite{translationdoc}.
These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.  These figures do not
include the spectacular speed-ups obtained in some cases by the JIT compiler
described in section \ref{subsect:dynamic_compilers}.

@@

%XXX doesn't look entirely nice
\begin{itemize}
\item \emph{Do not write dynamic language implementations ``by hand''.}
Writing them more abstractly, at a higher level, has primarily only
advantages, among them the avoidance of a proliferation of implementations
growing out of sync.  Writing interpreters both flexibly and efficiently
is difficult and meta-programming is a good way to achieve it.
Moreover, this is not incompatible with targeting and benefiting from
existing high-quality object-oriented virtual machines like those of the Java and .NET.

@@

The approach we outlined is actually just one in a very large, mostly
unexplored design space; it is likely that some of the choices made in
PyPy will turn out to be suboptimal.  We are hoping that other
toolchains will emerge over time, exploring other aspects and proposing
other solutions.  By their ``meta'' nature, these multiple approaches
should be easier to bridge together than, say, multiple OO VMs with
different object and runtime models.  We believe that further research
in this area might open the door to better solutions for
interoperability in general -- e.g. high-level bridges instead of
(virtual-)machine-level ones, enabled by cross-translation.
